Geometric models overestimate lake depth due to imperfect slope measurement

### 2020-06-09 manuscript/agujournaltemplate.tex
copy over to wrr template

---
title: Geometric models overestimate lake depth due to imperfect slope measurement

# Use letters for affiliations
author:
  - name: Joseph Stachelek
    affiliation: a
  - name: Patrick Hanly
    affiliation: a
  - name: Patricia A. Soranno
    affiliation: a

address:
  - code: a
    address: Department of Fisheries and Wildlife, Michigan State University, 480 Wilson Rd., East Lansing, MI 48824, USA

# Optional: line of arbitrary text with additional information.
# Could be used, for example, to mention the bibliographic info in a post-print.
# If not specified, defaults to "This version was compiled on \today"
#date_subtitle: Published in *Journal of Statistical Software*, 2018
# date_subtitle: > 
# This is the accepted version of the following article: J. Stachelek et al., Ecological Applications. In Press, which has been published in final form at DOI:  [DOI](https://doi.org).  This article may be used for non-commercial purposes in accordance with Wiley Terms and Conditions for Use of Self-Archived Versions.

# For footer text  TODO(fold into template, allow free form two-authors)
lead_author_surname: Stachelek et al.

# Place eg a DOI URL or CRAN Package URL here
# doi_footer: "https://doi.org/10.5281/zenodo.3754916"

# Abstract
abstract: |
  Lake depth is a critical characteristic that influences many important ecological processes in lakes. Unfortunately, lake depth measurements are labor-intensive to gather and are only available for a small fraction of lakes globally. Therefore, scientists have tried to predict lake depth from characteristics that are easily obtained for all lakes such as lake surface area or the slope of the land surrounding a lake. One approach for predicting lake depth simulates lake basins using a geometric model where nearshore land slope is assumed to be a representative proxy for in-lake slope and the distance to the center of the lake is assumed to be a representative proxy for the distance to the deepest point of the lake. However, these assumptions have rarely been tested in a broad range of lakes. We used bathymetry data from 5,000 lakes and reservoirs to test these assumptions and to examine whether differences in lake type or shape influences depth prediction error. We found that nearshore land slope was not a representative proxy of in-lake slope and using it for prediction increases prediction error substantially relative to models using true in-lake slope. We also found that models using nearshore land slope as a proxy systematically overpredicts lake depth in concave lakes (i.e. bowl-shaped; up to 18% of lakes in the study population) and reservoir lakes (up to 30% of lakes in the study population), suggesting caution in using geometric models for lake prediction in unsampled lakes.

# Optional: Acknowledgements
acknowledgements: |
  Funding was provided by the US NSF Macrosystems Biology Program grants, DEB-1638679; DEB-1638550, DEB-1638539, DEB-1638554. PAS was also supported by USDA National Institute of Food and Agriculture Hatch Project, Grant Number: 176820. Author contributions: JS conceived of the study, built models, analyzed data, and wrote the paper. PH and PAS provided interpretation of results and edited the paper.

# Optional: One or more keywords
keywords:
  - lake depth
  - bathymetry
  - hypsography
  - geometry
  - slope

# Paper size for the document, values of letter and a4
papersize: letter

# Font size of the document, values of 9pt (default), 10pt, 11pt and 12pt
fontsize: 9pt

# Optional: Force one-column layout, default is two-column
#one_column: true

# Optional: Enables lineno mode, but only if one_column mode is also true
#lineno: true

# Optional: Enable one-sided layout, default is two-sided
#one_sided: true

# Optional: Enable section numbering, default is unnumbered
#numbersections: true

# Optional: Specify the depth of section number, default is 5
#secnumdepth: 5

# Optional: Skip inserting final break between acknowledgements, default is false
skip_final_break: true

# Optional: Bibliography 
bibliography: lagosdepth
biblio-style: jsta

# Optional: Enable a 'Draft' watermark on the document
watermark: true

# Customize footer, eg by referencing the vignette
footer_contents: "Geometric lake depth models and bathymetry"

# Produce a pinp document
output: pinp::pinp

# Required: Vignette metadata for inclusion in a package.
vignette: >
  %\VignetteIndexEntry{YourPackage-vignetteentry}
  %\VignetteKeywords{YourPackage, r, anotherkeyword}
  %\VignettePackage{YourPackage}
  %\VignetteEngine{knitr::rmarkdown}
---

## INTRODUCTION

Lake depth is a major factor controlling lake physics, chemistry, and biota
\citep{liSignificantContributionLake2020, feeEffectsLakeSize1996, readSimulating2368Temperate2014}. However, because measured depth data is only available for a small fraction of all lakes, our ability to understand and predict depth-dependent processes is limited. The importance of lake depth, coupled with its limited availability, has led to numerous attempts to predict depth using measures available for all lakes such as lake surface area or the nearshore slope of the land surrounding a lake \citep{heathcotePredictingBathymetricFeatures2015, oliver2016prediction, sobekPredictingDepthVolume2011}. The approaches taken in these studies have been varied. In addition, they have explored a variety of strategies to maximize prediction accuracy such as employing more diverse covariates \citep{oliver2016prediction}, varying lake buffer sizes \citep{heathcotePredictingBathymetricFeatures2015}, or estimating hidden groupings among lakes \citep{sobekPredictingDepthVolume2011}.

One approach for predicting lake depth involves a geometric model that assumes lake basins correspond to an idealized shape such as a cone, bowl, or an elliptic sinusoid \citep{hollisterPredictingMaximumLake2011, neumannMaximumDepthAverage1959, yigzawNewGlobalStorage2018}. All such geometric models for lake depth prediction involve implicit assumptions about the terms of geometric formulae. In the simplest case, where lakes basins are treated as cones (Eq \ref{eq1}, Figure \ref{fig1}), two assumptions are required to make depth predictions for all lakes: 1) that nearshore land slope is a representative proxy for in-lake slope and 2) that the distance to the center of the lake is a representative proxy for the distance to the deepest point of the lake (Figure \ref{fig1}). This cone model imposes the following fixed (i.e. geometric) relationship between slope and horizontal distance:
\begin{align}
  depth_{geometric} = tan(slope) * distance \label{eq1}
\end{align}

where the product of slope and horizontal distance yields an exact geometric depth estimate (depth_geometric).

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=0.45\textwidth,keepaspectratio]{../figures/slope_diagram_new}
  \end{center}
  \caption{Diagram showing the relations between true (black) and proxy (orange) metrics of lake geometry. Geometric depth calculated via Equation \ref{eq1} requires a single distance and slope metric.}\label{fig1}
\end{figure}

The assumptions of this model as well as other geometric models can be tested by comparing proxy measures of lake geometry against corresponding “true” (i.e. in-lake) values derived from bathymetric maps. However, due to the scarcity of bathymetric map data, few studies have tested these assumptions except in individual studies of very large lakes or for small numbers of lakes (< 100) \citep{johanssonNewApproachesModelling2007}. As a result, it is unclear what effect proxies have on lake depth predictions using geometric models and whether depth predictions are more sensitive to errors in the horizontal dimension (i.e. distance to the deepest point of the lake) or errors in the vertical dimension (i.e. in-lake slope). Additionally it is unclear whether model prediction error is related to differences in lake type such those with different cross-section shapes or those classified as reservoirs versus natural lakes. This information could help direct future research efforts to focus on particular dimensions of lake geometry (i.e. horizontal versus vertical) or to stratify model predictions based on specific lake types and cross-section shapes.

@@

Given that, by definition, the distance proxy (distance to the center of the lake) must always be greater or equal to the true distance value (distance to the deepest point of the lake), we expect that the use of this proxy will lead to overestimation of lake depth (Figure \ref{fig1}). Furthermore we expect to see greater overestimation error in reservoirs as compared to natural lakes because many reservoirs are known to be drowned river valleys where the deepest point is close to the edge at the end of the reservoir (i.e. next to the dam) rather than in the center of the reservoir \citep{lanza1985interactions}. Similar to distance proxies, we expect that the use of a nearshore land slope proxy will lead to prediction error according to differences in cross-section shape (Figure \ref{fig2}) such that the depth of U-shaped (i.e. concave) lakes will be overpredicted whereas the depth of V-shaped (i.e. convex) lakes will be underpredicted. Finally, we expect that offsets to geometric depth estimates will be most strongly related to lake area and hydrologic subbasin variables as these measures have been influential in prior studies \citep{oliver2016prediction}.

\begin{figure}
  \begin{center}
    \includegraphics[width=0.45\textwidth,keepaspectratio]{../figures/lake_shape}
  \end{center}
  \caption{Diagram showing our expectation that slope-based models of lake depth will under predict true depth in convex lakes (left) and over predict true depth in concave lakes (right). Dashed lines represent extrapolated nearshore land slope while solid lines represent the lake bottom.}\label{fig2}
\end{figure}

## METHODS
### Data description
We compiled bathymetry data on approximately 5,000 lakes in the Northeastern and Midwestern US from nine official state databases (Figure S1). The original data came in a variety of formats including pre-interpolated rasters (Minnesota), contour lines (Nebraska, Michigan, Massachusetts, Kansas, Iowa), contour polygons (New Hampshire, Connecticut), or point depth soundings (Maine). For the Minnesota data, we simply clipped the raster for each lake to its outline. For data from the remaining states, we processed each lake by converting its original representation to a point layer (if necessary), rasterizing these points, and creating an interpolated bathymetry “surface” using a simple moving window average in the \texttt{raster} R package \citep{hijmansRasterGeographicData2019}. The size of the moving window was adjusted iteratively to ensure that each bathymetry raster contained no missing data.

All lake bathymetry was specifically calculated relative to high-resolution (1:24,000 scale) NHD \citep{usgsNationalHydrographyDataset2019} waterbodies such that source data and bathymetry surface outputs were clipped to the area of each lake polygon. We restricted the lakes in our study to those with an area of at least 4 ha and a maximum depth of at least 0.3 m (1 ft). The purpose of these restrictions was to ensure that lakes had enough contours (or points, or polygons) to generate adequately smooth interpolations with which to calculate in-lake geometry metrics.

We used our generated bathymetry surfaces to find the location of the deepest point in the lake and we resolved ties by choosing the deepest point that was closest to the center of the lake. We calculated the center of the lake not as its centroid but rather by finding the point farthest from the lake shoreline (i.e. its “visual distance to lake center”). For these calculations, we used the \texttt{polylabelr} R package \citep{larssonPolylabelrFindPole2019d}, which interfaces with the Mapbox pole of inaccessibility algorithm \citep{agafonkinJSLibraryFinding2019}. We calculated in-lake slope as maximum lake depth divided by the distance to the deepest point and we calculated nearshore land slope for each lake by computing the slope within a 100-m buffer using data from a high resolution digital elevation model (~15x15m grain) accessed using the \texttt{elevatr} R package \citep{hollisterElevatrAccessElevation2017} and the terrain function in the \texttt{raster} R package \citep{hijmansRasterGeographicData2019}.

We categorized lakes based on their cross-section shape and reservoir class. For cross-section shape, we categorized lakes as either convex or concave following the method of \citet{hakansonLakeFormLake1977} by computing normalized lake depth-area relationships (i.e. hypsographic curves) and assigning class membership based on whether a lake’s curve falls above or below that of a simple straight-sided cone (Figure S4). We further classified lakes using data from \citet{polus2020}, which uses the output of a machine learning algorithm to assign a probability to each lake regarding whether it is a reservoir or a natural lake. For our purposes, we determined a lake to be a reservoir if the classification probability was 0.75 or greater. The \citet{polus2020} data product defines reservoirs as any permanent waterbody that has a water control structure likely to significantly impact flow or pool water, beyond simply controlling water level. It makes no distinction between different dam types or dam heights.

\begin{table}
\centering
\setlength\tabcolsep{1.5pt} % default value: 6pt
\begin{tabular}{lllll}
  \hline
  Variable & Median & Q25 & Q75 & n\\
  \hline
  Max depth (m) & 8.2 (7) & 4.6 (3.7) & 14 (12) & 4850 (17700)\\
  \textbf{Elevation (m)} & 300 (340) & 180 (210) & 400 (460) & 4850 (17700)\\
  \textbf{Area (ha)} & 55 (33) & 21 (11) & 140 (100) & 4850 (17700)\\
  \textbf{Island area (ha)} & 0 (0) & 0 (0) & 0.18 (0.076) & 4850 (17700)\\
  \textbf{Perimeter (m)} & 4400 (3500) & 2500 (1800) & 8100 (7300) & 4850 (17700)\\
  \textbf{Shoreline development} & 1.7 (1.7) & 1.4 (1.4) & 2.1 (2.2) & 4850 (17700)\\
  \textbf{Watershed-lake ratio} & 7.8 (10) & 3.9 (4.4) & 17 (28) & 4850 (17700)\\
  Distance to deepest point (m) & 180 (-) & 110 (-) & 290 (-) & 4850 (-)\\
  Distance to lake center (m) & 240 (-) & 160 (-) & 380 (-) & 4850 (-)\\
  In-lake slope (m/m) & 0.046 (-) & 0.024 (-) & 0.079 (-) & 4850 (-)\\
  Nearshore land slope (m/m) & 0.077 (-) & 0.051 (-) & 0.11 (-) & 4850 (-)\\  
  \bottomrule
  \end{tabular}
  \caption{Summary of lake characteristics for the present study (and for lakes in the contiguous United States from <LAGOSUS-Depth citation>). Predictor variables for computing random forest offsets (Eq \ref{eq2}) are printed in bold face. Dashes (-) indicate an identical sample size among this study and that of the contiguous United States.} \label{table1}
\end{table}

@@

\begin{align}
shoreline_{devel} = perimeter / (2 * \sqrt{(\pi * waterarea * 10000)})       
\end{align}

where sinuous lakes have larger values of shoreline development and circular lakes have smaller values of shoreline development. Watershed to lake area ratio is an approximation of water residence time and is defined as watershed area divided by lake area. All of our data and data processing code are available at [Zenodo DOI].

### Proxy evaluation
We conducted a qualitative assessment of whether or not proxy measures of lake geometry are representative of their true values by visual inspection (i.e. plotting each proxy measure against its corresponding true value) and by computing coefficients of determination ($R^2$). We further tested proxy measures by examining their effect on lake depth prediction error. Our approach involved several steps. In the first step, we computed a geometric estimate of lake depth using only geometry information (depth_geometric, Equation \ref{eq1}). In the second step, we fit a random forest model to predict observed (i.e. true) depth as a function of geometric depth along with several covariates available for all lakes (Table \ref{table1}). The purpose of this random forest “offset” modeling was to more rigorously test our expectations regarding prediction error among different formulations of depth_geometric and among different lake types. Each of these steps were executed iteratively for each combination of true and proxy values of slope and distance (Table \ref{table2}).

## Model description
### Geometric model 
We used a geometric model of lakes where basins are treated as cones with a fixed relationship between slope and distance (Equation \ref{eq1}). Note that Equation \ref{eq1} is a geometric formula and has no intercept or “coefficients” and it produces a perfect depth estimate given true values of slope and distance. To use this model to predict the depth of all lakes, there is a necessary assumption that proxy slope and distance measures, which are available for all lakes, are representative of true slope and distance (Figure \ref{fig1}).

### Random forest models
Nearly all prior studies predicting lake depth using geometric models include a statistical or machine learning model “layer” or “offset” to boost predictive accuracy. For our purposes, this offset modeling enabled us to test our expectations that prediction error would be different among different formulations of depth_geometric and among different lake types. It also facilitated direct comparison against prior models of lake depth including those that are non-geometric. We generated an “offset” to geometric depth \citep[sensu][]{hollisterPredictingMaximumLake2011} using the random forest algorithm and the \texttt{ranger} R package \citep{wrightRangerFastImplementation2017} to predict observed maximum depth as a function of covariates including geometric maximum depth (from Equation \ref{eq1}) along with the lake elevation, area, perimeter, and ratio/index measures listed in Table 1:

\begin{align}
  depth_{observed} \sim depth_{geometric} + covariates         
\end{align}

Neither cross-section shape nor reservoir class was used as a covariate in any random forest models. We used the random forest algorithm because it makes no assumptions about the distribution of model residuals, allows for non-linearity, and is insensitive to interactions (i.e. multicollinearity) among covariates \citep{prasadNewerClassificationRegression2006}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% AGUJournalTemplate.tex: this template file is for articles formatted with LaTeX
%
% This file includes commands and instructions
% given in the order necessary to produce a final output that will
% satisfy AGU requirements, including customized APA reference formatting.
%
% You may copy this file and give it your
% article name, and enter your text.
%
%
% Step 1: Set the \documentclass
%
%

%% To submit your paper:
\documentclass[draft]{agujournal2019}
\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers
%%%%%%%
% As of 2018 we recommend use of the TrackChanges package to mark revisions.
% The trackchanges package adds five new LaTeX commands:
%
%  \note[editor]{The note}
%  \annote[editor]{Text to annotate}{The note}
%  \add[editor]{Text to add}
%  \remove[editor]{Text to remove}
%  \change[editor]{Text to remove}{Text to add}
%
% complete documentation is here: http://trackchanges.sourceforge.net/
%%%%%%%

\draftfalse

%% Enter journal name below.
%% Choose from this list of Journals:
%
% JGR: Atmospheres
% JGR: Biogeosciences
% JGR: Earth Surface
% JGR: Oceans
% JGR: Planets
% JGR: Solid Earth
% JGR: Space Physics
% Global Biogeochemical Cycles
% Geophysical Research Letters
% Paleoceanography and Paleoclimatology
% Radio Science
% Reviews of Geophysics
% Tectonics
% Space Weather
% Water Resources Research
% Geochemistry, Geophysics, Geosystems
% Journal of Advances in Modeling Earth Systems (JAMES)
% Earth's Future
% Earth and Space Science
% Geohealth
%
% ie, \journalname{Water Resources Research}

\journalname{Water Resources Research}

\begin{document}

%% ------------------------------------------------------------------------ %%
%  Title
%
% (A title should be specific, informative, and brief. Use
% abbreviations only if they are defined in the abstract. Titles that
% start with general keywords then specific terms are optimized in
% searches)
%
%% ------------------------------------------------------------------------ %%

% Example: \title{This is a test title}

\title{Geometric models overestimate lake depth due to imperfect slope
measurement}

%% ------------------------------------------------------------------------ %%
%
%  AUTHORS AND AFFILIATIONS
%
%% ------------------------------------------------------------------------ %%

% Authors are individuals who have significantly contributed to the
% research and preparation of the article. Group authors are allowed, if
% each author in the group is separately identified in an appendix.)

% List authors by first name or initial followed by last name and
% separated by commas. Use \affil{} to number affiliations, and
% \thanks{} for author notes.
% Additional author notes should be indicated with \thanks{} (for
% example, for current addresses).

% Example: \authors{A. B. Author\affil{1}\thanks{Current address, Antartica}, B. C. Author\affil{2,3}, and D. E.
% Author\affil{3,4}\thanks{Also funded by Monsanto.}}

\authors{J. Stachelek\affil{1}, P. Hanly\affil{1}, and P. A. Soranno\affil{1}}

\affiliation{1}{Department of Fisheries and Wildlife, Michigan State University, 480
Wilson Rd., East Lansing, MI 48824, USA}
% \affiliation{2}{Second Affiliation}
% \affiliation{3}{Third Affiliation}
% \affiliation{4}{Fourth Affiliation}

% \affiliation{=number=}{=Affiliation Address=}
%(repeat as many times as is necessary)

%% Corresponding Author:
% Corresponding author mailing address and e-mail address:

% (include name and email addresses of the corresponding author.  More
% than one corresponding author is allowed in this LaTeX file and for
% publication; but only one corresponding author is allowed in our
% editorial system.)

% Example: \correspondingauthor{First and Last Name}{email@address.edu}

\correspondingauthor{J. Stachelek}{stachel2@msu.edu}

%% Keypoints, final entry on title page.

%  List up to three key points (at least one is required)
%  Key Points summarize the main points and conclusions of the article
%  Each must be 140 characters or fewer with no special characters or punctuation and must be complete sentences

% Example:
% \begin{keypoints}
% \item	List up to three key points (at least one is required)
% \item	Key Points summarize the main points and conclusions of the article
% \item	Each must be 140 characters or fewer with no special characters or punctuation and must be complete sentences
% \end{keypoints}

\begin{keypoints}
\item Geometric models to predict lake depth, which require in-lake slope, assume that nearshore land slope is a good proxy for in-lake slope.
\item Using data from thousands of lakes, we show that nearshore land slope is a poor proxy for in-lake slope and increases prediction error.
\item Increases in prediction error were systematic such that depth was overpredicted in concave and reservoir lakes.
\end{keypoints}

%% ------------------------------------------------------------------------ %%
%
%  ABSTRACT and PLAIN LANGUAGE SUMMARY
%
% A good Abstract will begin with a short description of the problem
% being addressed, briefly describe the new data or analyses, then
% briefly states the main conclusion(s) and how they are supported and
% uncertainties.

% The Plain Language Summary should be written for a broad audience,
% including journalists and the science-interested public, that will not have 
% a background in your field.
%
% A Plain Language Summary is required in GRL, JGR: Planets, JGR: Biogeosciences,
% JGR: Oceans, G-Cubed, Reviews of Geophysics, and JAMES.
% see http://sharingscience.agu.org/creating-plain-language-summary/)
%
%% ------------------------------------------------------------------------ %%

%% \begin{abstract} starts the second page

\begin{abstract}
      Lake depth is a critical characteristic that influences many important
      ecological processes in lakes. Unfortunately, lake depth measurements
      are labor-intensive to gather and are only available for a small
      fraction of lakes globally. Therefore, scientists have tried to predict
      lake depth from characteristics that are easily obtained for all lakes
      such as lake surface area or the slope of the land surrounding a lake.
      One approach for predicting lake depth simulates lake basins using a
      geometric model where nearshore land slope is assumed to be a
      representative proxy for in-lake slope and the distance to the center of
      the lake is assumed to be a representative proxy for the distance to the
      deepest point of the lake. However, these assumptions have rarely been
      tested in a broad range of lakes. We used bathymetry data from 5,000
      lakes and reservoirs to test these assumptions and to examine whether
      differences in lake type or shape influences depth prediction error. We
      found that nearshore land slope was not a representative proxy of
      in-lake slope and using it for prediction increases prediction error
      substantially relative to models using true in-lake slope. We also found
      that models using nearshore land slope as a proxy systematically
      overpredicts lake depth in concave lakes (i.e.~bowl-shaped; up to 18\%
      of lakes in the study population) and reservoir lakes (up to 30\% of
      lakes in the study population), suggesting caution in using geometric
      models for lake prediction in unsampled lakes.
\end{abstract}

%% ------------------------------------------------------------------------ %%
%
%  TEXT
%
%% ------------------------------------------------------------------------ %%

%%% Suggested section heads:
% \section{Introduction}
%
% The main text should start with an introduction. Except for short
% manuscripts (such as comments and replies), the text should be divided
% into sections, each with its own heading.

% Headings should be sentence fragments and do not begin with a
% lowercase letter or number. Examples of good headings are:

% \section{Materials and Methods}
% Here is text on Materials and Methods.
%
% \subsection{A descriptive heading about methods}
% More about Methods.
%
% \section{Data} (Or section title might be a descriptive heading about data)
%
% \section{Results} (Or section title might be a descriptive heading about the
% results)
%
% \section{Conclusions}

\section{Introduction}
\noindent
Lake depth is a major factor controlling lake physics, chemistry, and biota
\cite{liSignificantContributionLake2020, feeEffectsLakeSize1996, readSimulating2368Temperate2014}. However, because measured depth data is only available for a small fraction of all lakes, our ability to understand and predict depth-dependent processes is limited. The importance of lake depth, coupled with its limited availability, has led to numerous attempts to predict depth using measures available for all lakes such as lake surface area or the nearshore slope of the land surrounding a lake \cite{heathcotePredictingBathymetricFeatures2015, oliver2016prediction, sobekPredictingDepthVolume2011}. The approaches taken in these studies have been varied. In addition, they have explored a variety of strategies to maximize prediction accuracy such as employing more diverse covariates \cite{oliver2016prediction}, varying lake buffer sizes \cite{heathcotePredictingBathymetricFeatures2015}, or estimating hidden groupings among lakes \cite{sobekPredictingDepthVolume2011}.

One approach for predicting lake depth involves a geometric model that assumes lake basins correspond to an idealized shape such as a cone, bowl, or an elliptic sinusoid \cite{hollisterPredictingMaximumLake2011, neumannMaximumDepthAverage1959, yigzawNewGlobalStorage2018}. All such geometric models for lake depth prediction involve implicit assumptions about the terms of geometric formulae. In the simplest case, where lakes basins are treated as cones (Eq \ref{eq1}, Figure \ref{fig1}), two assumptions are required to make depth predictions for all lakes: 1) that nearshore land slope is a representative proxy for in-lake slope and 2) that the distance to the center of the lake is a representative proxy for the distance to the deepest point of the lake (Figure \ref{fig1}). This cone model imposes the following fixed (i.e. geometric) relationship between slope and horizontal distance:

\begin{linenomath*}
\begin{equation}
      depth_{geometric} = tan(slope) * distance \label{eq1}
\end{equation}
\label{eq1}
\end{linenomath*}

\noindent
where the product of slope and horizontal distance yields an exact geometric depth estimate ($depth_{geometric}$).

\begin{figure}
\noindent\includegraphics[width=0.85\textwidth]{../figures/slope_diagram_new}
\caption{Diagram showing the relations between true (black) and proxy (orange) metrics of lake geometry. Geometric depth calculated via Equation \ref{eq1} requires a single distance and slope metric.}\label{fig1}
\end{figure}

The assumptions of this model as well as other geometric models can be tested by comparing proxy measures of lake geometry against corresponding “true” (i.e. in-lake) values derived from bathymetric maps. However, due to the scarcity of bathymetric map data, few studies have tested these assumptions except in individual studies of very large lakes or for small numbers of lakes (< 100) \cite{johanssonNewApproachesModelling2007}. As a result, it is unclear what effect proxies have on lake depth predictions using geometric models and whether depth predictions are more sensitive to errors in the horizontal dimension (i.e. distance to the deepest point of the lake) or errors in the vertical dimension (i.e. in-lake slope). Additionally it is unclear whether model prediction error is related to differences in lake type such those with different cross-section shapes or those classified as reservoirs versus natural lakes. This information could help direct future research efforts to focus on particular dimensions of lake geometry (i.e. horizontal versus vertical) or to stratify model predictions based on specific lake types and cross-section shapes.

@@

Given that, by definition, the distance proxy (distance to the center of the lake) must always be greater or equal to the true distance value (distance to the deepest point of the lake), we expect that the use of this proxy will lead to overestimation of lake depth (Figure \ref{fig1}). Furthermore we expect to see greater overestimation error in reservoirs as compared to natural lakes because many reservoirs are known to be drowned river valleys where the deepest point is close to the edge at the end of the reservoir (i.e. next to the dam) rather than in the center of the reservoir \cite{lanza1985interactions}. Similar to distance proxies, we expect that the use of a nearshore land slope proxy will lead to prediction error according to differences in cross-section shape (Figure \ref{fig2}) such that the depth of U-shaped (i.e. concave) lakes will be overpredicted whereas the depth of V-shaped (i.e. convex) lakes will be underpredicted. Finally, we expect that offsets to geometric depth estimates will be most strongly related to lake area and hydrologic subbasin variables as these measures have been influential in prior studies \cite{oliver2016prediction}.

\begin{figure}
      \begin{center}
        \includegraphics[width=0.75\textwidth,keepaspectratio]{../figures/lake_shape}
      \end{center}
      \caption{Diagram showing our expectation that slope-based models of lake depth will under predict true depth in convex lakes (left) and over predict true depth in concave lakes (right). Dashed lines represent extrapolated nearshore land slope while solid lines represent the lake bottom.}\label{fig2}
    \end{figure}

\section{Methods}
\subsection{Data description}
\noindent
We compiled bathymetry data on approximately 5,000 lakes in the Northeastern and Midwestern US from nine official state databases (Figure S1). The original data came in a variety of formats including pre-interpolated rasters (Minnesota), contour lines (Nebraska, Michigan, Massachusetts, Kansas, Iowa), contour polygons (New Hampshire, Connecticut), or point depth soundings (Maine). For the Minnesota data, we simply clipped the raster for each lake to its outline. For data from the remaining states, we processed each lake by converting its original representation to a point layer (if necessary), rasterizing these points, and creating an interpolated bathymetry “surface” using a simple moving window average in the \texttt{raster} R package \cite{hijmansRasterGeographicData2019}. The size of the moving window was adjusted iteratively to ensure that each bathymetry raster contained no missing data.

All lake bathymetry was specifically calculated relative to high-resolution (1:24,000 scale) NHD \cite{usgsNationalHydrographyDataset2019} waterbodies such that source data and bathymetry surface outputs were clipped to the area of each lake polygon. We restricted the lakes in our study to those with an area of at least 4 ha and a maximum depth of at least 0.3 m (1 ft). The purpose of these restrictions was to ensure that lakes had enough contours (or points, or polygons) to generate adequately smooth interpolations with which to calculate in-lake geometry metrics.

We used our generated bathymetry surfaces to find the location of the deepest point in the lake and we resolved ties by choosing the deepest point that was closest to the center of the lake. We calculated the center of the lake not as its centroid but rather by finding the point farthest from the lake shoreline (i.e. its “visual distance to lake center”). For these calculations, we used the \texttt{polylabelr} R package \cite{larssonPolylabelrFindPole2019d}, which interfaces with the Mapbox pole of inaccessibility algorithm \cite{agafonkinJSLibraryFinding2019}. We calculated in-lake slope as maximum lake depth divided by the distance to the deepest point and we calculated nearshore land slope for each lake by computing the slope within a 100-m buffer using data from a high resolution digital elevation model (~15x15m grain) accessed using the \texttt{elevatr} R package \cite{hollisterElevatrAccessElevation2017} and the terrain function in the \texttt{raster} R package \cite{hijmansRasterGeographicData2019}.

We categorized lakes based on their cross-section shape and reservoir class. For cross-section shape, we categorized lakes as either convex or concave following the method of \citeA{hakansonLakeFormLake1977} by computing normalized lake depth-area relationships (i.e. hypsographic curves) and assigning class membership based on whether a lake’s curve falls above or below that of a simple straight-sided cone (Figure S4). We further classified lakes using data from \citeA{polus2020}, which uses the output of a machine learning algorithm to assign a probability to each lake regarding whether it is a reservoir or a natural lake. For our purposes, we determined a lake to be a reservoir if the classification probability was 0.75 or greater. The \citeA{polus2020} data product defines reservoirs as any permanent waterbody that has a water control structure likely to significantly impact flow or pool water, beyond simply controlling water level. It makes no distinction between different dam types or dam heights.

\begin{table}
\caption{Summary of lake characteristics for the present study (and for lakes in the contiguous United States from <LAGOSUS-Depth citation>). Predictor variables for computing random forest offsets (Eq \ref{eq2}) are printed in bold face. Dashes (-) indicate an identical sample size among this study and that of the contiguous United States.} \label{table1}
\centering
\setlength\tabcolsep{1.5pt} % default value: 6pt
\begin{tabular}{lllll}
  \hline
  Variable & Median & Q25 & Q75 & n\\
  \hline
  Max depth (m) & 8.2 (7) & 4.6 (3.7) & 14 (12) & 4850 (17700)\\
  \textbf{Elevation (m)} & 300 (340) & 180 (210) & 400 (460) & 4850 (17700)\\
  \textbf{Area (ha)} & 55 (33) & 21 (11) & 140 (100) & 4850 (17700)\\
  \textbf{Island area (ha)} & 0 (0) & 0 (0) & 0.18 (0.076) & 4850 (17700)\\
  \textbf{Perimeter (m)} & 4400 (3500) & 2500 (1800) & 8100 (7300) & 4850 (17700)\\
  \textbf{Shoreline development} & 1.7 (1.7) & 1.4 (1.4) & 2.1 (2.2) & 4850 (17700)\\
  \textbf{Watershed-lake ratio} & 7.8 (10) & 3.9 (4.4) & 17 (28) & 4850 (17700)\\
  Distance to deepest point (m) & 180 (-) & 110 (-) & 290 (-) & 4850 (-)\\
  Distance to lake center (m) & 240 (-) & 160 (-) & 380 (-) & 4850 (-)\\
  In-lake slope (m/m) & 0.046 (-) & 0.024 (-) & 0.079 (-) & 4850 (-)\\
  Nearshore land slope (m/m) & 0.077 (-) & 0.051 (-) & 0.11 (-) & 4850 (-)\\  
  \hline    
  \end{tabular}  
\end{table}

@@

\begin{linenomath*}
      \begin{equation}
            shoreline_{devel} = perimeter / (2 * \sqrt{(\pi * waterarea * 10000)}) \label{eq2}        
      \end{equation}
      \end{linenomath*}

\noindent
where sinuous lakes have larger values of shoreline development and circular lakes have smaller values of shoreline development. Watershed to lake area ratio is an approximation of water residence time and is defined as watershed area divided by lake area. All of our data and data processing code are available at [Zenodo DOI].

\subsubsection{Proxy evaluation}
\noindent
We conducted a qualitative assessment of whether or not proxy measures of lake geometry are representative of their true values by visual inspection (i.e. plotting each proxy measure against its corresponding true value) and by computing coefficients of determination ($R^2$). We further tested proxy measures by examining their effect on lake depth prediction error. Our approach involved several steps. In the first step, we computed a geometric estimate of lake depth using only geometry information ($depth_{geometric}$, Equation \ref{eq1}). In the second step, we fit a random forest model to predict observed (i.e. true) depth as a function of geometric depth along with several covariates available for all lakes (Table \ref{table1}). The purpose of this random forest “offset” modeling was to more rigorously test our expectations regarding prediction error among different formulations of $depth_{geometric}$ and among different lake types. Each of these steps were executed iteratively for each combination of true and proxy values of slope and distance (Table \ref{table2}).

\subsection{Model description}
\subsubsection{Geometric model}
\noindent
We used a geometric model of lakes where basins are treated as cones with a fixed relationship between slope and distance (Equation \ref{eq1}). Note that Equation \ref{eq1} is a geometric formula and has no intercept or “coefficients” and it produces a perfect depth estimate given true values of slope and distance. To use this model to predict the depth of all lakes, there is a necessary assumption that proxy slope and distance measures, which are available for all lakes, are representative of true slope and distance (Figure \ref{fig1}).

\subsubsection{Random forest models}
\noindent
Nearly all prior studies predicting lake depth using geometric models include a statistical or machine learning model “layer” or “offset” to boost predictive accuracy. For our purposes, this offset modeling enabled us to test our expectations that prediction error would be different among different formulations of $depth_{geometric}$ and among different lake types. It also facilitated direct comparison against prior models of lake depth including those that are non-geometric. We generated an “offset” to geometric depth \cite<sensu>{hollisterPredictingMaximumLake2011} using the random forest algorithm and the \texttt{ranger} R package \cite{wrightRangerFastImplementation2017} to predict observed maximum depth as a function of covariates including geometric maximum depth (from Equation \ref{eq1}) along with the lake elevation, area, perimeter, and ratio/index measures listed in Table 1:

\begin{linenomath*}
      \begin{equation}
            depth_{observed} \sim depth_{geometric} + covariates \label{eq3}
      \end{equation}
      \end{linenomath*}

\noindent      
Neither cross-section shape nor reservoir class was used as a covariate in any random forest models. We used the random forest algorithm because it makes no assumptions about the distribution of model residuals, allows for non-linearity, and is insensitive to interactions (i.e. multicollinearity) among covariates \cite{prasadNewerClassificationRegression2006}.

\subsubsection{Model comparisons}
\noindent
We tested model sensitivity to slope and distance proxies by generating multiple “geometric maximum depth” estimates from 3 different model runs using each of the possible metric combinations for Equation (true slope - proxy distance, proxy slope - true distance, proxy slope - proxy distance). Prior to entry into Equation \ref{eq1}, we standardized proxy distances to have the same numeric range as their true counterpart. The purpose of this standardization was to prevent lakes with extremely long proxy distances from having an outsized impact on model evaluation metrics.
